# Computational Statistics Lab 5 
Authors: Christian Kammerer, Jakob Lindner

## Question 1:

a. Read in the dataset and fit a cubic regression model $y = β_0 + β_1x + β_2x^2 + β_3x^3 + \epsilon$, where x = concentration and y = yield, using the R-function lm.

```{r}
set.seed(42)
data <- read.table("kresseertrag.dat", header=FALSE)

model <- lm(V3~V2+I(V2^2)+I(V2^3), data=data)
summary(model)
```

b. You might improve the model in a. by removing or adding (a) model term(s). For the chosen model, estimate the coefficients in the model together with their 95%-confidence intervals using the R-function lm. Create a plot for yield vs. concentration and add the estimated regression curve to the plot.

The summary of the fitted model shows, that the term $\beta_1x$ has a very low significance. To simplify the model we remove the term which results in the following model: $y = β_0 + β_2x^2 + β_3x^3 + \epsilon$ 

Using the confint function for lm we get the confidence interval for the parameters.

```{r}
model <- lm(V3~I(V2^2)+I(V2^3), data=data)
summary(model)

model$coefficients
confint.lm(model, level=0.95)

```

```{r echo=FALSE}
plot(data$V2, data$V3, main="Regression fitted using lm", ylab="yield", xlab="concentration of fertilizer")
x <- seq(0,1.2,0.01)
df_x <- data.frame(V2=x)
y <- predict(model, newdata=df_x)
lines(x,y, col="red", lwd=2)
```


c. Derive a 95%-bootstrap confidence interval for one of the model parameters βi based on the percentile method. Do not use a bootstrap package for this calculation; program the bootstrap on your own. Use at least 10000 bootstrap replicates. Plot a histogram with the bootstrap distribution for the parameter.

For this task we adjusted the code from the lecture to train a model on the sampled data in each iteration. The values for $\beta_2$ are saved and plotted in the histogram.

```{r}
bo <- 10000 
bs <- numeric(length(bo)) 
for (l in 1:bo){
  x  <- data[sample(nrow(data), replace=TRUE), ]
  m <- lm(V3~I(V2^2)+I(V2^3), data=x) 
  bs <- c(bs, m$coefficients[2]) # getting coefficient for beta2
}

```

```{r echo=FALSE}
hist(bs)
bss <- sort(bs)
ci95 <- c(bss[round(bo*0.025)], bss[round(bo*0.975)])
cat("95%-confidence interval using percentile method: ", ci95, "\n")
```


d. Derive now 95%-confidence intervals for the chosen parameter using the package boot with percentile and BCa method.

```{r echo=FALSE}
library(boot)
```

The "boot"-package offers the functionality to get confidence intervals using the percentile and the BCa method.

```{r}
fert <- function(data, i){
  m <- lm(V3~I(V2^2)+I(V2^3), subset=i,data=data)
  m$coefficients[2]
}

cb <- boot(data, fert, R=10000)

ci_percentile <- boot.ci(cb, type = "perc")
ci_bca <- boot.ci(cb, type = "bca")
ci_percentile
ci_bca
```

e. Compare the confidence intervals from b., c., and d. and comment on it. Do you judge that the intervals are similar or do you see relevant differences? What is your overall conclusion from these confidence interval results about that model parameter?

The confidence intervals retrieved from bootstrapping in c. and d. are very similar around (-226, -90). With (-231.83687 -84.86522) the CI calculated using confint.lm() is a bit lower.  

Confint assumes normality according to the documentation, while bootstrapping does not require to make any assumptions on the distribution. But the confidence interval for the parameter retrieved with confint is still close to the result from bootstrapping and not a relevent difference. Therefore, the parameter is almost normally distributed. The shape of the plotted histogram above confirms this statement.  

## Appendix

```{r eval=FALSE}
set.seed(42)
data <- read.table("kresseertrag.dat", header=FALSE)

model <- lm(V3~V2+I(V2^2)+I(V2^3), data=data)
summary(model)


model <- lm(V3~I(V2^2)+I(V2^3), data=data)
summary(model)

model$coefficients
confint.lm(model, level=0.95)

plot(data$V2, data$V3, main="Regression fitted using lm", ylab="yield", xlab="concentration of fertilizer")
x <- seq(0,1.2,0.01)
df_x <- data.frame(V2=x)
y <- predict(model, newdata=df_x)
lines(x,y, col="red", lwd=2)



# adjusted code from lecture slides

bo <- 10000 
bs <- numeric(length(bo)) 
for (l in 1:bo){
  x  <- data[sample(nrow(data), replace=TRUE), ]
  m <- lm(V3~I(V2^2)+I(V2^3), data=x) 
  bs <- c(bs, m$coefficients[2]) # getting coefficient for beta2
}


hist(bs)
bss <- sort(bs)
ci95 <- c(bss[round(bo*0.025)], bss[round(bo*0.975)])
cat("95%-confidence interval using percentile method: ", ci95, "\n")


library(boot)

fert <- function(data, i){
  m <- lm(V3~I(V2^2)+I(V2^3), subset=i,data=data)
  m$coefficients[2]
}

cb <- boot(data, fert, R=10000)

ci_percentile <- boot.ci(cb, type = "perc")
ci_bca <- boot.ci(cb, type = "bca")
ci_percentile
ci_bca
``` 